---
title: "Model-Quality Experiment — Football Iceland 2026"
subtitle: "Can an optimal Bayesian model + Kelly process profit from Lengjan's static line?"
format:
  dashboard:
    theme: cosmo
    embed-resources: true
    orientation: rows
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(gt)
  library(jsonlite)
})

dash_dir <- here::here("data", "dashboard")
files <- list.files(dash_dir, pattern = "\\.json$", full.names = TRUE)
D <- lapply(files, function(f) jsonlite::fromJSON(f))
names(D) <- tools::file_path_sans_ext(basename(files))

base_theme <- tryCatch(metill::theme_metill(), error = function(e) theme_minimal(base_size = 12))
pct <- function(x, d = 1) sprintf(paste0("%+.", d, "f%%"), 100 * x)
pct0 <- function(x) sprintf("%.0f%%", 100 * x)

meta <- D$meta
sk <- D$skill
mb <- D$market_bias
ls_ <- D$line_stability
dr_sex <- dplyr::filter(D$draw_rate, scope == "sex")

total_sk <- sk[sk$market == "total", , drop = FALSE]
ml_home <- mb[mb$market == "moneyline" & mb$outcome == "home", , drop = FALSE]
verdict_color <- if (nrow(total_sk) && total_sk$brier_skill[1] < 0) "danger" else "success"
```

# Status

## Row {height="22%"}

```{r}
#| content: valuebox
#| title: "Model vs. de-vigged market — TOTALS"
list(
  value = if (nrow(total_sk)) pct(total_sk$brier_skill[1]) else "n/a",
  icon = "graph-down-arrow",
  color = verdict_color
)
```

```{r}
#| content: valuebox
#| title: "OOS matches scored"
list(value = meta$n_pit_total[1], icon = "calendar-check", color = "secondary")
```

```{r}
#| content: valuebox
#| title: "Lengjan price moves (share of series)"
list(value = if (nrow(ls_)) pct0(ls_$pct_moved[1]) else "n/a", icon = "arrow-left-right", color = "info")
```

```{r}
#| content: valuebox
#| title: "Draw gap (obs − pred, male)"
list(
  value = {
    g <- dr_sex$gap[dr_sex$sex == "male"]
    if (length(g)) pct(g[1]) else "n/a"
  },
  icon = "dribbble", color = "secondary"
)
```

## Row

::: {.card title="The experiment, in one screen"}

**Hypothesis chain:** an optimal Bayesian model (is it *right*?) → a soft, static
line (where is it *wrong*?) → the model is *sharper* than the line → +EV → a
self-updating Kelly/bankroll → realised profit. The leading indicators below
accumulate over every match; realised PnL is the slow lagging confirmation.

**What the diagnostics say so far (male, `r meta$n_pit_total[1]` OOS matches):**

- **Totals are the model's blind spot** — the de-vigged line out-forecasts the
  model on over/under (Brier skill **`r if (nrow(total_sk)) pct(total_sk$brier_skill[1]) else "n/a"`**),
  the only market where the CI excludes zero. On *spread* the model is competitive.
- **The line is soft on home & overs** — it under-prices home wins and overs and
  over-prices draws (see *Where is the line soft?*).
- **The line is not frozen** — Lengjan moves ~`r if (nrow(ls_)) pct0(ls_$pct_moved[1]) else "?"`
  of price series at least once, so closing-line value is an open empirical
  question, not a dead end.

*Female market diagnostics are pending a `team_names` name-join fix in the
walk-forward (`bt_wf_require_pre_cutoff_odds`); female PIT / draw-rate are shown.*

:::

# Is the model right?

## Row

```{r}
#| title: "Randomised PIT — total goals (uniform = calibrated shape)"
pt <- D$pit_total
if (is.data.frame(pt) && nrow(pt)) {
  ggplot(pt, aes(u)) +
    geom_histogram(boundary = 0, bins = 10, fill = "#2c7fb8", colour = "white") +
    geom_hline(yintercept = nrow(pt) / 10 / length(unique(pt$sex)), linetype = 2, colour = "grey40") +
    facet_wrap(~sex, scales = "free_y") +
    labs(x = "PIT value", y = "matches") +
    base_theme
} else {
  ggplot() +
    annotate("text", 0, 0, label = "no PIT data") +
    theme_void()
}
```

```{r}
#| title: "Randomised PIT — goal difference (Skellam)"
pd <- D$pit_diff
if (is.data.frame(pd) && nrow(pd)) {
  ggplot(pd, aes(u)) +
    geom_histogram(boundary = 0, bins = 10, fill = "#7fcdbb", colour = "white") +
    facet_wrap(~sex, scales = "free_y") +
    labs(x = "PIT value", y = "matches") +
    base_theme
} else {
  ggplot() +
    theme_void()
}
```

## Row

```{r}
#| title: "Predicted vs observed draw rate (the draw-deficit check)"
dr <- dplyr::filter(D$draw_rate, scope == "division")
if (is.data.frame(dr) && nrow(dr)) {
  dr |>
    dplyr::arrange(sex, division) |>
    dplyr::transmute(sex, division, n,
      predicted = predicted_draw_rate, observed = observed_draw_rate, gap
    ) |>
    gt() |>
    fmt_percent(c(predicted, observed, gap), decimals = 1) |>
    data_color(columns = gap, palette = c("#d73027", "white", "#1a9850"), domain = c(-0.2, 0.2)) |>
    tab_options(table.font.size = px(12))
} else {
  gt(tibble::tibble(note = "no draw-rate data"))
}
```

```{r}
#| title: "Scoreline residual grid (observed − predicted; off-diagonal = correlation fault)"
sr <- D$scoreline
if (is.data.frame(sr) && nrow(sr)) {
  ggplot(sr, aes(factor(home_goals), factor(away_goals), fill = residual)) +
    geom_tile(colour = "white") +
    scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9850", midpoint = 0) +
    facet_wrap(~sex) +
    labs(x = "home goals", y = "away goals", fill = "obs−pred") +
    base_theme
} else {
  ggplot() +
    theme_void()
}
```

# Where is the line soft?

## Row

```{r}
#| title: "Market bias per outcome — bias > 0 means the line UNDER-prices it (value pocket)"
if (is.data.frame(mb) && nrow(mb)) {
  mb |>
    dplyr::arrange(sex, market, outcome) |>
    dplyr::transmute(sex, market, outcome, n, line_implied = mean_q, realised, bias) |>
    gt() |>
    fmt_percent(c(line_implied, realised, bias), decimals = 1) |>
    data_color(columns = bias, palette = c("#d73027", "white", "#1a9850"), domain = c(-0.12, 0.12)) |>
    tab_options(table.font.size = px(12))
} else {
  gt(tibble::tibble(note = "no market-bias data"))
}
```

## Row

```{r}
#| title: "Disagreement — where model and line diverge, who is right?"
dg <- D$disagreement
if (is.data.frame(dg) && nrow(dg)) {
  ord <- c("model<<mkt", "model<mkt", "agree", "model>mkt", "model>>mkt")
  dg |>
    dplyr::mutate(band = factor(band, levels = ord)) |>
    dplyr::arrange(sex, market, band) |>
    dplyr::transmute(sex, market, band, n, model_p = mean_p, line_q = mean_q, realised) |>
    gt() |>
    fmt_percent(c(model_p, line_q, realised), decimals = 1) |>
    tab_options(table.font.size = px(11))
} else {
  gt(tibble::tibble(note = "no disagreement data"))
}
```

```{r}
#| title: "Does Lengjan move prices?"
if (is.data.frame(ls_) && nrow(ls_)) {
  ls_ |>
    dplyr::transmute(
      `price series` = n_series,
      `share that moved` = pct0(pct_moved),
      `mean distinct prices` = round(mean_distinct_prices, 2),
      `mean snapshots` = round(mean_snapshots, 2)
    ) |>
    gt() |>
    tab_options(table.font.size = px(13))
} else {
  gt(tibble::tibble(note = "no odds data"))
}
```

# Am I sharper than the line?

## Row

```{r}
#| title: "Model vs. de-vigged market — Brier skill by market (90% match-clustered CI)"
if (is.data.frame(sk) && nrow(sk)) {
  sk |>
    dplyr::arrange(sex, market) |>
    dplyr::transmute(sex, market, n,
      matches = n_matches,
      brier_skill, ci = sprintf("[%+.1f, %+.1f]%%", 100 * skill_lo, 100 * skill_hi)
    ) |>
    gt() |>
    fmt_percent(brier_skill, decimals = 1) |>
    data_color(columns = brier_skill, palette = c("#d73027", "white", "#1a9850"), domain = c(-0.1, 0.1)) |>
    tab_options(table.font.size = px(12))
} else {
  gt(tibble::tibble(note = "no skill data"))
}
```

## Row

```{r}
#| title: "Murphy decomposition — compare reliability & resolution across markets (not raw Brier)"
bd <- D$brier_decomp
if (is.data.frame(bd) && nrow(bd)) {
  bd |>
    dplyr::arrange(sex, market) |>
    dplyr::transmute(sex, market, n,
      reliability = round(reliability, 4), resolution = round(resolution, 4),
      uncertainty = round(uncertainty, 4), brier = round(brier, 4)
    ) |>
    gt() |>
    tab_options(table.font.size = px(12)) |>
    tab_footnote("Lower reliability = better calibrated; higher resolution = better discrimination.")
} else {
  gt(tibble::tibble(note = "no decomposition data"))
}
```
