---
title: "Betting Backtest — Iceland 2026"
subtitle: "What would have been working, and how well"
author: "Brynjólfur Gauti Jónsson"
date: today
date-format: "YYYY-MM-DD"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 144
    embed-resources: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
invisible(Sys.setlocale("LC_ALL", "is_IS.UTF-8"))
options(width = 110)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(gt)
  library(scales)
  devtools::load_all(here::here(), quiet = TRUE)
})

results <- read_table("results")
initial_pool <- load_bankroll()$initial_pool
picks <- bt_load_universe(strategy = "kept")
roll <- bt_run(picks, results, stake_rule = stake_rolling, initial_pool = initial_pool)
fixed <- bt_run(picks, results, stake_rule = stake_fixed, initial_pool = initial_pool)
```

## Headline — our actual picks

```{r}
#| label: kpis
m <- bt_metrics(roll)
gt(m) |>
  fmt_number(c(total_staked, total_pnl, max_drawdown), decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Rolling-bankroll performance (kept bets)")
```

```{r}
#| label: pnl-curve
#| fig-alt: "Cumulative profit-and-loss of the bankroll over the backtest window."
roll |>
  arrange(match_date) |>
  ggplot(aes(match_date, pool_after)) +
  geom_hline(yintercept = initial_pool, linetype = 2, colour = "grey50") +
  geom_step() +
  labs(
    x = NULL, y = "Bankroll (ISK)",
    title = "Rolling bankroll — kept bets"
  ) +
  theme_minimal()
```

## Where the edge is

```{r}
#| label: by-dimension
bind_rows(
  bt_metrics(fixed, by = "market") |> rename(dim = market) |> mutate(facet = "market"),
  bt_metrics(fixed, by = "sport") |> rename(dim = sport) |> mutate(facet = "sport"),
  bt_metrics(fixed, by = "sex") |> rename(dim = sex) |> mutate(facet = "sex")
) |>
  select(facet, dim, n_bets, total_pnl, roi, hit_rate, avg_odds) |>
  gt(groupname_col = "facet") |>
  fmt_number(total_pnl, decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Fixed-fraction ROI by dimension")
```

## Calibration

```{r}
#| label: calibration
#| fig-alt: "Reliability plot: predicted probability versus realised win frequency."
bt_calibration(fixed, n_bins = 8) |>
  filter(n > 0) |>
  ggplot(aes(mean_p, realised_freq, size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.7) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Predicted p", y = "Realised frequency",
    title = "Calibration (kept bets)"
  ) +
  theme_minimal()
```

## Strategy comparison

```{r}
#| label: baselines
bt_baselines() |>
  select(strategy, stake_model, n_bets, total_pnl, roi, hit_rate) |>
  arrange(desc(roi)) |>
  gt() |>
  fmt_number(total_pnl, decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Strategy x stake-model comparison")
```

## Caveats

- **Leak-free** picks: `candidates.p` was frozen pre-match; settlement reuses
  `compute_settlement()` so win/push boundaries match the live decider.
- **Obtainability:** odds are decide-time prices; some bets were never
  placeable (early-kickoff delisting). Pending/unsettled bets are excluded.
- **Counterfactual stakes are approximate** for non-kept strategies
  (`portfolio_lambda` / calibration are not stored in `candidates`).
- **Window** is short (candidates begin 2026-04-25) — per-cell slices with low
  `n_bets` are noisy.
