---
title: "Betting Backtest — Football Iceland 2026"
subtitle: "How well football betting would have worked, and in which leagues"
author: "Brynjólfur Gauti Jónsson"
date: today
date-format: "YYYY-MM-DD"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    fig-dpi: 144
    embed-resources: true
    smooth-scroll: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
invisible(Sys.setlocale("LC_ALL", "is_IS.UTF-8"))
options(width = 110)
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(gt)
  library(scales)
  devtools::load_all(here::here(), quiet = TRUE)
})

results <- read_table("results")
initial_pool <- load_bankroll()$initial_pool
# Football only -- this report informs football specifically; basketball and
# handball are on seasonal pause (and would dilute the football verdict).
picks <- bt_load_universe(strategy = "kept", leagues = "football")
roll <- bt_run(picks, results, stake_rule = stake_rolling, initial_pool = initial_pool)
fixed <- bt_run(picks, results, stake_rule = stake_fixed, initial_pool = initial_pool)

# Per-league (division) view. The betting stores carry no division; recover it
# from results on the federation-name match key so every breakdown names the
# Icelandic league. Bettable football is men's BD/LD1/LD2/CUP + women's BD; men's
# 3. deild and women's Lengjudeild/2. deild are modelled but never priced by
# Lengjan, so they cannot appear in a betting backtest (see Caveats).
roll <- bt_attach_division(roll, results)
fixed <- bt_attach_division(fixed, results)

# Canonical division code -> Icelandic label, per sex, from config/leagues.yml
# (publish_divisions); falls back to the raw code for anything unmapped.
div_levels <- c("BD", "LD1", "LD2", "LD3", "LD4", "CUP", "unknown")
league_label <- function(sex, division) {
  out <- division
  for (s in unique(sex)) {
    m <- tryCatch(.football_iceland_division_labels(s), error = function(e) character())
    i <- sex == s & division %in% names(m)
    out[i] <- unname(m[division[i]])
  }
  out
}
```

```{r}
#| label: wf-setup
#| include: false

# Walk-forward (out-of-sample) arm. REUSE mode reconstructs each saved fit's
# predictions (beliefs/extracts/predicted_matches) and re-decides -- no Stan, a
# whole season in seconds -- so this report self-updates on every render as new
# fits land. Empty (gitignored extracts absent) -> the WF sections skip cleanly.
wf_sexes <- c("male", "female")
wf_sexes <- wf_sexes[dir.exists(here::here(
  "data", "beliefs", "extracts", "sport=football",
  "country=iceland", paste0("sex=", wf_sexes)
))]
wf_bets <- dplyr::bind_rows(lapply(wf_sexes, function(s) {
  r <- bt_walkforward_reuse(s, season = 2026)
  if (is.null(r)) NULL else r$bets
}))
# Attach division before devig so the model-vs-market arm slices by league too.
if (nrow(wf_bets) > 0) wf_bets <- bt_attach_division(wf_bets, results)
wf_available <- isTRUE(nrow(wf_bets) > 0)

# De-vigged model-vs-market table (margin-free market probability `q_market`),
# restricted to complete, settled, non-push (match, market, line) books.
mkt <- if (wf_available) bt_devig(wf_bets) else tibble::tibble()
```

```{r}
#| label: wf-empty-note
#| output: asis
#| echo: false
if (!wf_available) {
  cat(
    "::: {.callout-note}\n",
    "No walk-forward output yet. Generate it with ",
    "`Rscript scripts/0Nb_walkforward.R --sex male --reuse` ",
    "(seconds), then re-render.\n:::\n"
  )
}
```

## Headline — our actual picks

```{r}
#| label: kpis
m <- bt_metrics(roll)
gt(m) |>
  fmt_number(c(total_staked, total_pnl, max_drawdown), decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Rolling-bankroll performance (kept bets)")
```

```{r}
#| label: pnl-curve
#| fig-alt: "Cumulative profit-and-loss of the bankroll over the backtest window."
roll |>
  arrange(match_date) |>
  ggplot(aes(match_date, pool_after)) +
  geom_hline(yintercept = initial_pool, linetype = 2, colour = "grey50") +
  geom_step() +
  labs(
    x = NULL, y = "Bankroll (ISK)",
    title = "Rolling bankroll — kept bets"
  ) +
  theme_minimal()
```

## Leagues covered

```{r}
#| label: coverage-note
#| output: asis
cov <- fixed |>
  dplyr::count(sex, division) |>
  dplyr::mutate(
    division = factor(division, div_levels),
    lab = paste0(league_label(sex, as.character(division)), " (", n, ")")
  ) |>
  dplyr::arrange(sex, division)
cov_line <- function(s) {
  v <- cov$lab[cov$sex == s]
  if (length(v) == 0) "_none_" else paste(v, collapse = ", ")
}
cat(
  "::: {.callout-note}\n",
  "## Every bettable Icelandic football league this season\n\n",
  "**Men:** ", cov_line("male"), "  \n",
  "**Women:** ", cov_line("female"), "\n\n",
  "Bracketed numbers are kept bets (fixed-era, deduped). Men's *3. deild* and ",
  "women's *Lengjudeild*/*2. deild* are modelled but Lengjan never prices them, ",
  "so they cannot enter a betting backtest — absent by necessity, not omission.\n",
  ":::\n",
  sep = ""
)
```

## Where the edge is

```{r}
#| label: by-dimension
# Football only, so sport is constant -- break down by the dimensions that vary
# within football: market type and sex.
bind_rows(
  bt_metrics(fixed, by = "market") |> rename(dim = market) |> mutate(facet = "market"),
  bt_metrics(fixed, by = "sex") |> rename(dim = sex) |> mutate(facet = "sex")
) |>
  select(facet, dim, n_bets, total_pnl, roi, hit_rate, avg_odds) |>
  gt(groupname_col = "facet") |>
  fmt_number(total_pnl, decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Fixed-fraction ROI by dimension (football)")
```

## By league

Every bettable league broken out. Men's **Besta deild / Lengjudeild / 2. deild /
Mjólkurbikar** and women's **Besta deild** are the cells Lengjan priced this
season. Per-league `n` is small, so read these as directional, not significant.

```{r}
#| label: by-league
bt_metrics(fixed, by = c("sex", "division")) |>
  mutate(
    league = league_label(sex, division),
    division = factor(division, div_levels)
  ) |>
  arrange(sex, division) |>
  select(sex, league, n_bets, total_pnl, roi, hit_rate, avg_odds) |>
  gt(groupname_col = "sex") |>
  fmt_number(total_pnl, decimals = 0) |>
  fmt_number(avg_odds, decimals = 2) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  cols_label(
    league = "League", n_bets = "Bets", total_pnl = "PnL (ISK)",
    roi = "ROI", hit_rate = "Hit rate", avg_odds = "Avg odds"
  ) |>
  tab_header("Fixed-fraction performance by league (kept bets)") |>
  tab_source_note("Per-league n is small — directional, not significant.")
```

## Calibration

```{r}
#| label: calibration
#| fig-alt: "Reliability plot: predicted probability versus realised win frequency."
bt_calibration(fixed, n_bins = 8) |>
  filter(n > 0) |>
  ggplot(aes(mean_p, realised_freq, size = n)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.7) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Predicted p", y = "Realised frequency",
    title = "Calibration (kept bets)"
  ) +
  theme_minimal()
```

## Walk-forward — does the model hold up out of sample?

The sections above judge our *actual bets*. These judge the *model*: every row
below is a prediction made for a round the model had **not** been fit on, with
the out-of-sample window capped at the next fit so nothing is double-counted or
leaked. This is the whole-season picture (all bettable candidates, not just the
handful we placed), so n is far larger than the kept-bet calibration above.

```{r}
#| label: wf-scores
#| eval: !expr wf_available
wf_bets |>
  dplyr::group_by(sex) |>
  dplyr::group_modify(~ bt_oos_scores(.x)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    brier_skill = 1 - brier / 0.25,
    logloss_skill = 1 - log_loss / log(2)
  ) |>
  gt::gt(rowname_col = "sex") |>
  gt::fmt_number(c(brier, log_loss), decimals = 3) |>
  gt::fmt_percent(c(brier_skill, logloss_skill), decimals = 1) |>
  gt::cols_label(
    n = "OOS preds", brier = "Brier", log_loss = "Log-loss",
    brier_skill = "Brier skill", logloss_skill = "Log-loss skill"
  ) |>
  gt::tab_header("Out-of-sample probability scores (vs. coin-flip)") |>
  gt::tab_source_note("Skill > 0 beats a 50/50 guess. 0.25 / 0.693 are the coin-flip Brier / log-loss.")
```

```{r}
#| label: wf-reliability
#| eval: !expr wf_available
#| fig-alt: "Out-of-sample reliability: model predicted probability versus realised win frequency per probability bin, with binomial intervals, faceted by sex."
wf_bets |>
  dplyr::filter(!is.na(win), is.finite(p)) |>
  dplyr::mutate(y = as.numeric(win), bin = cut(p, seq(0, 1, 0.1), include.lowest = TRUE)) |>
  dplyr::group_by(sex, bin) |>
  dplyr::summarise(
    mean_p = mean(p), realised = mean(y), n = dplyr::n(),
    lo = qbeta(.05, sum(y) + .5, dplyr::n() - sum(y) + .5),
    hi = qbeta(.95, sum(y) + .5, dplyr::n() - sum(y) + .5),
    .groups = "drop"
  ) |>
  dplyr::filter(n > 0) |>
  ggplot(aes(mean_p, realised)) +
  geom_abline(slope = 1, linetype = 2, colour = "grey55") +
  geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey75") +
  geom_point(aes(size = n)) +
  facet_wrap(~sex) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Model predicted p", y = "Realised frequency",
    title = "Walk-forward reliability (all OOS predictions)"
  ) +
  theme_minimal()
```

## Model vs. market — is there an informational edge?

The bookmaker's odds are a forecast too. After stripping the vig (normalising
`1/odds` to sum to 1 within each match-market book), whoever has the lower Brier
is the better forecaster on these matches. A positive **Brier skill score**
means the model carries information the line did not.

```{r}
#| label: wf-edge-readout
#| eval: !expr wf_available
#| output: asis
#| echo: false
edge <- bt_skill(mkt)
ci <- bt_skill_ci(mkt)
verdict <- dplyr::case_when(
  ci[1] > 0 ~ "a **credible edge** over the market",
  ci[3] < 0 ~ "the market is **credibly sharper** than the model",
  TRUE ~ "**no demonstrable edge** either way — indistinguishable from the de-vigged line"
)
cat(sprintf(
  paste0(
    "> Over **%d** out-of-sample predictions across **%d** matches, the model's ",
    "Brier skill vs. the de-vigged Lengjan line is **%+.1f%%** ",
    "(90%% match-clustered CI %+.1f%% to %+.1f%%) — %s.\n"
  ),
  nrow(mkt), edge$n_matches, 100 * edge$brier_skill, 100 * ci[1], 100 * ci[3], verdict
))
```

```{r}
#| label: wf-vs-market
#| eval: !expr wf_available
dplyr::bind_rows(
  dplyr::mutate(bt_skill(mkt), market = "ALL"),
  bt_skill(mkt, by = "market")
) |>
  dplyr::relocate(market) |>
  gt::gt(rowname_col = "market") |>
  gt::fmt_number(c(brier_model, brier_market, logloss_model, logloss_market), decimals = 3) |>
  gt::fmt_percent(c(brier_skill, logloss_skill, overround), decimals = 1) |>
  gt::cols_label(
    n = "n preds", n_matches = "matches",
    brier_model = "model", brier_market = "market",
    logloss_model = "model", logloss_market = "market",
    brier_skill = "Brier skill", logloss_skill = "LogL skill", overround = "Overround"
  ) |>
  gt::tab_spanner("Brier", c(brier_model, brier_market)) |>
  gt::tab_spanner("Log-loss", c(logloss_model, logloss_market)) |>
  gt::tab_header("Model vs. de-vigged market — OOS prediction quality") |>
  gt::tab_source_note("Skill > 0 = model beats the margin-free line on that market. Overround = bookmaker margin.")
```

The same comparison, sliced by league. The model is fit jointly across all
divisions, so this asks whether its informational edge over the line concentrates
in any particular league.

```{r}
#| label: wf-vs-market-league
#| eval: !expr wf_available
dplyr::bind_rows(
  dplyr::mutate(bt_skill(mkt, by = "sex"), division = "ALL"),
  bt_skill(mkt, by = c("sex", "division"))
) |>
  dplyr::mutate(
    league = ifelse(division == "ALL", "— all —", league_label(sex, division)),
    division = factor(division, c("ALL", div_levels))
  ) |>
  dplyr::arrange(sex, division) |>
  dplyr::select(sex, league, n, n_matches, brier_model, brier_market, brier_skill) |>
  gt::gt(groupname_col = "sex", rowname_col = "league") |>
  gt::fmt_number(c(brier_model, brier_market), decimals = 3) |>
  gt::fmt_percent(brier_skill, decimals = 1) |>
  gt::cols_label(
    n = "n preds", n_matches = "matches",
    brier_model = "model", brier_market = "market", brier_skill = "Brier skill"
  ) |>
  gt::tab_spanner("Brier", c(brier_model, brier_market)) |>
  gt::tab_header("Model vs. de-vigged market — by league") |>
  gt::tab_source_note("Brier skill > 0 = model beats the margin-free line in that league. Per-league n is thin — directional only.")
```

```{r}
#| label: wf-reliability-overlay
#| eval: !expr wf_available
#| fig-height: 4.2
#| fig-alt: "Reliability curves for the model and the de-vigged market overlaid on the same axes, faceted by market type."
dplyr::bind_rows(
  dplyr::transmute(mkt, market, source = "Model", prob = p, y),
  dplyr::transmute(mkt, market, source = "Market", prob = q_market, y)
) |>
  dplyr::mutate(bin = cut(prob, seq(0, 1, 0.1), include.lowest = TRUE)) |>
  dplyr::group_by(market, source, bin) |>
  dplyr::summarise(mean_p = mean(prob), realised = mean(y), n = dplyr::n(), .groups = "drop") |>
  dplyr::filter(n >= 5) |>
  ggplot(aes(mean_p, realised, colour = source)) +
  geom_abline(slope = 1, linetype = 2, colour = "grey55") +
  geom_line(alpha = .5) +
  geom_point(aes(size = n)) +
  facet_wrap(~market) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_colour_manual(values = c(Model = "#2c3e50", Market = "#e67e22")) +
  labs(
    x = "Predicted p", y = "Realised frequency", colour = NULL,
    title = "Model vs. market reliability, by market"
  ) +
  theme_minimal()
```

When the model disagrees with the line, does it pay? Bets only happen on
disagreement, so this is the betting-relevant cut: in the *model > market* rows,
if realised frequency tracks the model's p rather than the market's, the
disagreement carries information.

```{r}
#| label: wf-disagreement
#| eval: !expr wf_available
mkt |>
  dplyr::mutate(
    gap = p - q_market,
    band = cut(gap, c(-Inf, -.05, -.01, .01, .05, Inf),
      labels = c("model << mkt", "model < mkt", "~ agree", "model > mkt", "model >> mkt")
    )
  ) |>
  dplyr::group_by(band) |>
  dplyr::summarise(
    n = dplyr::n(), model_p = mean(p), market_q = mean(q_market),
    realised = mean(y), .groups = "drop"
  ) |>
  gt::gt(rowname_col = "band") |>
  gt::fmt_percent(c(model_p, market_q, realised), decimals = 1) |>
  gt::cols_label(model_p = "Model p", market_q = "Market q", realised = "Realised") |>
  gt::tab_header("When the model disagrees with the line, who is right?") |>
  gt::tab_source_note("Where model > market: if Realised tracks Model p (not Market q), the model's disagreements carry information.")
```

## Strategy comparison

```{r}
#| label: baselines
bt_baselines(leagues = "football") |>
  select(strategy, stake_model, n_bets, total_pnl, roi, hit_rate) |>
  arrange(desc(roi)) |>
  gt() |>
  fmt_number(total_pnl, decimals = 0) |>
  fmt_percent(c(roi, hit_rate), decimals = 1) |>
  tab_header("Strategy x stake-model comparison (football)")
```

## Caveats

- **Football only.** Basketball and handball are excluded (seasonal pause, and
  this report is meant to judge football specifically). Re-include all sports
  with `Rscript scripts/0Nb_backtest.R --league all` once they resume.
- **Bettable leagues only.** Both arms run through the decider, which needs a
  Lengjan price, so they cover men's *Besta deild / Lengjudeild / 2. deild /
  Mjólkurbikar* and women's *Besta deild*. Men's *3. deild* and women's
  *Lengjudeild*/*2. deild* are fit by the same model but never quoted by Lengjan,
  so a betting backtest cannot include them — their absence is structural, not a
  data gap. Division is recovered from `results` on the federation-name match key;
  fixtures with no result yet show as `unknown` (none among settled kept bets).
- **Leak-free** picks: `candidates.p` was frozen pre-match; settlement reuses
  `compute_settlement()` so win/push boundaries match the live decider.
- **Obtainability:** odds are decide-time prices; some bets were never
  placeable (early-kickoff delisting). Pending/unsettled bets are excluded.
- **Counterfactual stakes are approximate** for non-kept strategies
  (`portfolio_lambda` / calibration are not stored in `candidates`).
- **Window** is short (candidates begin 2026-04-25) — per-cell slices with low
  `n_bets` are noisy.
- **Walk-forward is leak-free by construction.** Each OOS prediction comes from a
  fit that never saw its match; the window is capped at the next fit so no match
  is double-counted. Reconstructed from the saved per-fit predictions
  (`beliefs/extracts/`) — no re-fit, so the report self-updates on every render.
- **Market = de-vigged line.** `1/odds` is normalised to sum to 1 within each
  `(match, market, line)` book, removing the bookmaker margin so model and market
  are scored on the same scale. Comparison runs only on complete, settled,
  non-push books; proportional de-vig mildly overstates favourites (a small tilt
  in the model's favour).
- **Model vs. market is selection-conditioned, not "model beats the bookie."**
  It scores only the matches Lengjan priced (and that cleared the candidate
  filter): "is the model competitive with the line on the bets we'd consider?",
  not a claim about fixtures the market declined to price.
- **Correlated within a match.** The three moneyline outcomes (and the
  spread/total sides) of one fixture are mechanically dependent, so the
  Brier-skill CI is bootstrapped by **match**, not by row — effective n is well
  below the prediction count. At a single season's scale, read by-market skill
  scores as directional, not significant.
